---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SelectionBias

<!-- badges: start -->
<!-- badges: end -->

The goal of SelectionBias is to calculate different bounds for the
selection bias for binary outcome and treatment variables. The sensitivity 
parameters for the SV (Smith and VanderWeele) bound and the GAF (generalized
assumption-free) bound are calculated for the generalized M-structure based on
user input parameters. The SV and GAF bounds can then be calculated either based
on these or other sensitivity parameter given by the. The CAF (counterfactual
assumption-free) bound is calculated based on sensitivity parameters and a 
dataset/probabilites from data given by the user. The AF (assumption-free) bound
is calculated for a dataset/probabilities from data.

## Installation

You can install the development version of SelectionBias from Github
(https://github.com/stizet/SelectionBias) with:

``` r
# install.packages("devtools") 
# devtools::install_github("stizet/SelectionBias")
```

## Bound examples

Selections of the study population can be the source of bias. Here, different
types of bounds for the selection bias is calculated; the bound by Smith and
VanderWeele (2019) and the GAF bounds, which depends on untestable assumptions,
and the CAF and AF bounds.

The simulated dataset `zika_learner` is included for the purpose of
illustration of the bounds. The simulated dataset is created to emulate real
data as well as a previous example. See references in corresponding article
and the data documentation.

The zika data can be reached through
```{r}
library(SelectionBias)
data(zika_learner)
attach(zika_learner)
```

The sensitivity parameters in the SV and GAF bounds are calculated in 
`sensitivityparametersM()` for user defined parameters in the extended
M-structure, where causal dependencies are modeled with either probit or logit
models. Which sensitivity parameters are indicated by the argument `whichBound`.
Below is an example with binary unobserved variables and two selection variables
(zika example in corresponding article):
```{r}
# library(SelectionBias)
V = matrix(c(1, 0, 0.85, 0.15), nrow = 2)
U = matrix(c(1, 0, 0.5, 0.5), nrow = 2)
Tr = c(-6.2, 1.75)
Y = c(-5.2, 5, -1)
S = matrix(c(1.2, 2.2, 0, 0.5, 2, -2.75, -4, 0), ncol = 4)

# SV bound.
sensitivityparametersM(whichEst = "RR_sub", whichBound = "SV", Vval = V,
                       Uval = U, Tcoef = Tr, Ycoef = Y, Scoef = S, Mmodel = "L",
                       pY1_T1_S1 = 0.286, pY1_T0_S1 = 0.004)

# GAF bound.
sensitivityparametersM(whichEst = "RR_sub", whichBound = "GAF", Vval = V,
                       Uval = U, Tcoef = Tr, Ycoef = Y, Scoef = S, Mmodel = "L",
                       pY1_T1_S1 = 0.286, pY1_T0_S1 = 0.004)
```
The output for the SV bound is the sensitivity parameters and an indicator that
states if the treatment was recoded, i.e. that the bias in the original coding
was negative. (See original article for details.) The output for the GAF bound
is the sensitivity parameters. No recoding of the treatment is necessary since
both an upper and a lower GAF bound is defined.

The SV bound is calculated in `SVbound()` where the user input is the
sensitivity parameters. These can either be found for the extended M-structure 
in `sensitivityparametersM()`, or from another source. Below is an example where
the input is the output from `sensitivityparametersM()` above:
```{r}
# library(SelectionBias)
SVbound(whichEst = "RR_sub", pY1_T1_S1 = 0.004, pY1_T0_S1 = 0.286,
        RR_UY_S1 = 2.71, RR_TU_S1 = 2.33)
```
The output is the SV bound. Here, the treatment has been recoded as indicated by
`sensitivityparametersM()`. Note that the eventual recoding of the treatment has
to be done manually, as discussed in the corresponding article.

The GAF bounds are calculated in `GAFbound()` where the user input is the
sensitivity parameters and either a dataset or probabilities from the data. The
sensitivity parameters can either be found for the extended M-structure 
in `sensitivityparametersM()`, or from another source. If interest lies in the
total population, the selection variable or selection probability must be given.
If interest lies in the subpopulation, only data on the selected individuals 
must be given. Below is an example where the input is the output from
`sensitivityparametersM()` above and the `zika_learner` dataset:
```{r}
# library(SelectionBias)
GAFbound(whichEst = "RR_sub", M = 0.4502, m = 0.002, 
         outcome = mic_ceph[sel_ind == 1], treatment = zika[sel_ind == 1])
```
The output is the GAF lower and upper bounds. Note that the treatment does not
need to be recoded since there is both a lower and upper bound. Please have this
in mind when comparing the bounds to the SV bound.

The CAF bounds are calculated in `CAFbound()` where the user input is the
sensitivity parameters and either a dataset or probabilities from the data. If
interest lies in the total population, the selection variable or selection
probability must be given. If interest lies in the subpopulation, only data on
the selected individuals must be given. Below is an example where the input is
sensitivity parameters given by the user and the conditional outcome and
treatment probabilities (see the documentation for more details) from the
`zika_learner`
dataset:
```{r}
# library(SelectionBias)
CAFbound(whichEst = "RR_sub", M = 0.5, m = 0.001, outcome = c(0.286, 0.004),
          treatment = c(0.002, 0.998))
```
The output is the CAF lower and upper bounds. Note that the treatment does not
need to be recoded since there is both a lower and upper bound. Please have this
in mind when comparing the bounds to the SV bound.

The AF bound is calculated in `AFbound()` where the user input is either a
dataset or probabilities from the data. If interest lies in the total
population, the selection variable or selection probability must be given. If 
interest lies in the subpopulation, only data on the selected individuals must 
be given.
```{r}
# library(SelectionBias)
AFbound(whichEst = "RR_sub", outcome = mic_ceph[sel_ind == 1],
        treatment = zika[sel_ind == 1])
AFbound(whichEst = "RR_sub", outcome = c(0.286, 0.004),
        treatment = c(0.002, 0.998))
```
The output is the AF lower and upper bounds. The difference in these two
examples comes from rounding errors. Note that the treatment does not need to be
recoded since there is both a lower and upper bound. Please have this in mind
when comparing the bounds to the SV bound.


## Sharp example

The sharpness of the SV bound in the subpopulation can be assessed in the 
function `SVboundsharp()`. The input is $BF_U$, $P(Y=1|T=0,I_S=1)$ and the SV
and AF bounds. Below is an example where the input is the output from
previous functions:
```{r}
# library(SelectionBias)
SVboundsharp(BF_U = 1.56, pY1_T0_S1 = 0.27)
```
The output states if the bound is sharp, or if it is inconclusive. There are
two optional arguments as well, SVbound and AFbound. These must be entered if
one wish to know if the bound is *not* sharp.
```{r}
# library(SelectionBias)
SVboundsharp(BF_U = 1.56, pY1_T0_S1 = 0.27, SVbound = 1.56, AFbound = 3.5)
```
The output states if the bound is sharp, inconclusive or not sharp. Note that
the eventual recoding of the treatment has to be done manually, as discussed
in the corresponding article.

